Adaptive Video Streaming with Optimized Bitstream Extraction and PID-based Quality Control

### 2015-05-11 tmm.tex
change the order of the two parts

% paper title
% can use linebreaks \\ within to get better formatting as desired
% Do not put math or special symbols in the title.
\title{Adaptive Video Streaming with PID-Based Quality Control and Optimized Bitstream Extraction}

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
To cope with the challenge brought by bandwidth fluctuation and provide the user with video of both smooth and high quality, an adaptive video streaming framework is proposed based on the Scalable Video Coding (SVC) extension of H.264/AVC. First, the Proportional-Integral-Derivative (PID) method is utilized to control the video quality adjustment according to actual bandwidth condition. By monitoring and predicting bandwidth information of the past, the current and the future, the PID-based quality control algorithm determines a suitable bitrate for transmission. Then, under the constraint of that bitrate, an efficient scheme is used to extract an optimal sub-stream from the whole SVC video bitstream. In this scheme, a linear error model which exploits the linear feature of pixel value errors is used to accurately estimate the distortion caused by discarding any combination of data packets, and a greedy-like algorithm is designed to assign each data packet a priority value according to its Rate-Distortion (R-D) impact, thus enables R-D optimized bitstream extraction. Experimental results show that, comparing with the baseline software, the proposed framework integrating the above algorithms can achieve much lower video quality fluctuation, with PSNR variance reduced from 0.66 to 0.28, and at the same time deliver higher video quality, with an average PSNR gain of 0.33dB. The framework is deployed at online video website www.7dlive.com and proves to be practically effective.
\end{abstract}

@@

One of the biggest challenge for video streaming is the variety of users' network environment. A user may use different types of network (Ethernet, Wi-Fi, Cellular, etc.) and may also switch from one type to another (e.g., from Wi-Fi to cellular network) on going. Even in the same network, the user's bandwidth may change from time to time, influenced by the overall traffic. During such an environment, bandwidth fluctuation is inevitable. If the video is streamed at a constant bitrate, it would be difficult or impossible for the user to get best watching experience. To cope with this challenge, the video streaming system should be able to automatically adjust the video's bitrate or quality according to the bandwidth change. In other words, the video streaming system needs to be adaptive.

There are generally two options to enable adaptive video streaming. The first option is called simulcast, where several independent bitstreams with different video quality are stored at the server, and the client selects one of them according to currently available bandwidth. Dynamic Adaptive Streaming over HTTP (DASH) \cite{DASH}, which is standardized by the Moving Picture Expert Group (MPEG), belongs to this option. The second option is to adopt the Scalable Video Coding (SVC) \cite{SVC} technology, an extension of the H.264/AVC video coding standard \cite{SVCOverview}, which allows scalability of video bitrate and quality with only one single bitstream. In SVC, a video bitstream is composed of small data packets called Network Abstract Layer Units (NALUs). These NALUs are separated into one base layer for basic video quality and multiple enhancement layers for video quality improvement. By discarding NALUs of the enhancement layers, the bitstream can be extracted to provide the most appropriate data rate and video quality as required.

The advantage of simulcast systems is that they need little modification of existing infrastructure and therefore are easy to deploy. Most commercial video websites, YouTube \cite{YouTube} for instance, use this option to enable quality adaptation. However, the adaptive range and granularity that simulcast provides is limited to the number of streams (YouTube videos have 5 quality levels at most: 240p, 360p, 480p, 720p and 1080p), and the processing and storage of multiple video streams significantly increase time and space cost. SVC streaming systems, on the contrary, do not have above shortcomings. Since an SVC video bitstream can be extracted at the granularity of one NALU (a very small partition of data), it is able to provide seamless scalability and quality adaptation. Besides, performance analysis \cite{SVCPerformance} shows that SVC also has much higher coding efficiency than simulcast, despite some computation and bitrate overhead comparing with single layer stream. Therefore, we choose to adopt SVC in our adaptive video streaming system. While the work of this paper can be partly used in a simulcast system, it will mainly focus on the adaptive streaming system using SVC.

In the SVC video streaming system, one scalable video stream with enhancement data packets is produced according to actual requirement and stored at the server for streaming (how to configure and produce the stream is not in the scope of this paper; users are referred to \cite{JSVM}). During the streaming process, there are two key problems that need to be addressed well. The first is quality control, and the second is bitstream extraction. Quality control means controlling the video quality (or the video bitrate, from the transmission point of view) to follow the change of the bandwidth. Once the bandwidth has changed, the quality control algorithm should determine a bitrate suitable for the current bandwidth. Given this bitrate, how to extract data from the whole SVC stream to form an optimal sub-stream becomes the second problem, i.e., bitstream extraction problem. Optimized bitstream extraction aims to select a subset of NALUs (or packets) which represents video with best quality, or minimal distortion, under the constraint of the given bitrate. It should be noted that, bitstream extraction is not only useful in adaptive video streaming, but also in other SVC application scenarios where part of data needs to be obtained or kept, e.g., the archiving of surveillance video.

In this paper, an efficient adaptive video streaming system based on SVC is proposed with the above two problems better solved. Apple Darwin Streaming Server (DSS) \cite{DSS}, which is open-sourced and has been widely used to stream multimedia contents in practice, is chosen as the basis of the system. The original DSS project is modified by us and SVC support is added to it, for the purpose of both research and application. Then novel algorithms of quality control and bitstream extraction are designed and implemented to achieve significant improvement in terms of user-perceived video quality. The contributions of our work are threefold:
\begin{itemize}
\item A quality control algorithm based on the Proportional-Integral-Derivative (PID) control concept is designed to replace the native one in DSS, and achieves higher average video quality with less quality fluctuation.
\item A linear error model is proposed to accurately estimate the distortion caused by discarding any combination of packets from a SVC bitstream, based on which a greedy-like bitstream extraction method is proposed and achieves PSNR gain of up to 0.5dB.
\item The proposed novel solutions for quality control and bitstream extraction are combined together and result in a higher and smoother PSNR curve representing the final video quality perceived by the user.
\end{itemize}

The rest of this paper is organized as follows. In section \ref{sec:analysis}, we analyze the problems in detail and introduce some related works on them. Then in section \ref{sec:quality-control}, the proposed quality control algorithm is formulated and explained with term definitions and parameters selection. After that, we present the solution for the bitstream extraction problem in section \ref{sec:bit-extraction}, including the linear error model and the greed-like algorithm. Section \ref{sec:experiment} lists the experimental results and compares our proposed solution with others. Finally, in section \ref{sec:conclusion}, we conclude this paper's work and point out its promising applications and extensions in the future.

@@

The two most concerned problems in SVC adaptive streaming, i.e., quality control and bitstream extraction, are illustrated in Fig. \ref{fig:problems}. In this section, we analyze the two problems in detail, and introduce some existing related works on them.

\begin{figure}[t]
\centering
\includegraphics[width = 0.9\linewidth]{Problems.jpg}
\caption{Two key problems in adaptive video streaming\label{fig:problems}}
\end{figure}

@@

The following of this paper will present solutions for the above problems in detail. We have published the related ideas in conference papers \cite{Meng14}\cite{Meng14-PID}, and in this paper we combine our previous works together and lay a solid foundation for a practical adaptive video streaming system.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Optimized Bitstream Extraction Based on Linear Error Model}
\label{sec:bit-extraction}

After the quality control algorithm determines an appropriate bitrate and sets a quality level accordingly, the system will only send packets belonging to this new quality level (as described at the end of Algorithm \ref{algo:control}). For SVC videos, it means to extract a subset of packets from the whole bitstream. How to optimally choose this subset to achieve minimum distortion under the given bitrate is the key point of bitstream extraction problem. In this section, we propose a linear error mode and a greedy-like algorithm to solve this problem.

We have presented experimental results for the case where reference original sequence is available in our previous work \cite{Zhang12}, so this paper will include only experimental results for the no-reference case in the following section.

@@

To implement our proposed adaptive streaming system requires the combination of the PID-based quality control algorithm and the optimized bitstream extraction method. The former determines the current bitrate to follow the bandwidth change, and the latter will ensure the sub-stream with the best possible quality is obtained and sent under that bitrate constraint. They work together to construct a system that delivers best quality of experience for the users. Choosing DSS as the implementation basis, we have integrated the proposed algorithms into it for the purpose of both research and application. In this section, we first present separate experiments to demonstrate the advantage of each proposed solution, and then give the result when they are combined together. Readers who are interested in the practical implementation can also visit the online video website http://www.7dlive.com for more information.

>>>>

% paper title
% can use linebreaks \\ within to get better formatting as desired
% Do not put math or special symbols in the title.
\title{Adaptive Video Streaming with Optimized Bitstream Extraction and PID-based Quality Control}

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
To cope with the challenge brought by bandwidth fluctuation and provide the user with video of both smooth and high quality, an adaptive video streaming framework which can adjust video quality according to actual bandwidth condition is proposed based on the Scalable Video Coding (SVC) extension of H.264/AVC. First, a simple and effective linear error model is proposed and validated for quality scalability of SVC. The model exploits the linear feature of pixel value errors and can be used to accurately estimate the distortion caused by discarding any combination of enhancement data packets in an SVC bitstream. Then on the basis of that, a greedy-like algorithm is designed to assign each data packet a priority value according to its Rate-Distortion (R-D) impact, thus enables R-D optimized bitstream extraction under certain bitrate constraint. Finally, the Proportional-Integral-Derivative (PID) method is utilized to control the video quality adjustment and determine the suitable bitrate for transmission. By monitoring and predicting bandwidth information of the past, the current and the future, the PID-based quality control algorithm can significantly reduce quality fluctuation while still keeping a high quality level. Experimental results show that, comparing with the baseline software, the proposed framework integrating the above algorithms can achieve much lower video quality fluctuation, with PSNR variance reduced from 0.66 to 0.28, and at the same time deliver higher video quality, with an average PSNR gain of 0.33dB. The framework is deployed at online video website www.7dlive.com and proves to be practically effective.
\end{abstract}

@@

One of the biggest challenge for video streaming is the variety of users' network environment. A user might use different types of network (Ethernet, Wi-Fi, Cellular, etc.) and might also switch from one type to another (e.g., from Wi-Fi to cellular network) on going. Even in the same network, the user's bandwidth will change from time to time, influenced by the overall traffic. Within such an environment, bandwidth fluctuation is inevitable. If the video is streamed at a constant bitrate, it would be difficult or impossible for the user to get best watching experience. To cope with this challenge, the video streaming system should be able to automatically adjust the video's bitrate or quality according to the bandwidth change. In other words, the video streaming system needs to be adaptive.

There are generally two options to enable adaptive video streaming. The first option is called simulcast, where several independent bitstreams each with different video quality are stored at the server, and the client selects one of them according to currently available bandwidth. Dynamic Adaptive Streaming over HTTP (DASH) \cite{DASH}, which is standardized by the Moving Picture Expert Group (MPEG), belongs to this option. The second option is to adopt the Scalable Video Coding (SVC) \cite{SVC} technology, an extension of the H.264/AVC video coding standard \cite{SVCOverview}, which allows scalability of video bitrate and quality with only one single bitstream. In SVC, a video bitstream is composed of small data packets called Network Abstract Layer Units (NALUs). These NALUs are separated into one base layer for basic video quality and multiple enhancement layers for video quality improvement. By discarding NALUs of the enhancement layers, the bitstream can be extracted to provide the most appropriate data rate and video quality as required.

The advantage of simulcast systems is that they need little modification of existing infrastructure and therefore are easy to deploy. Most commercial video websites, YouTube \cite{YouTube} for instance, use this option to enable quality adaptation. However, the adaptive range and granularity that simulcast provides is limited to the number of streams (YouTube videos have 5 quality levels at most: 240p, 360p, 480p, 720p and 1080p), and the processing and storage of multiple video streams significantly increase time and space cost. SVC streaming systems, on the contrary, do not have above shortcomings. Since an SVC video bitstream can be extracted at the granularity of one NALU (a very small partition of data), it is able to provide seamless scalability and quality adaptation. Besides, performance analysis \cite{SVCPerformance} shows that SVC also has much higher coding efficiency than simulcast, despite some computation and bitrate overhead comparing with single layer stream. Therefore, we choose to adopt SVC in our adaptive video streaming framework. While the work of this paper partly applies to the simulcast system, it will mainly focus on the context of SVC video streaming system.

In the SVC adaptive video streaming system, there are two key problems that need to be addressed well. First, as the prerequisite of the adaption to bandwidth change, the whole bitstream needs to be extracted so that sub-streams with various bitrates can be obtained. This is called the bitstream extraction problem. Optimal bitstream extraction algorithm aims to extract a sub-stream with best video quality, or minimal distortion, under the constraint of a given bitrate. Second, when the bandwidth does change, it's necessary to make decisions about whether and how to adjust the bitrate or video quality. This problem is called quality control, which means controlling the video quality (or the video bitrate, from the transmission point of view) to follow the change of the bandwidth. A good quality control algorithm should determine a suitable bitrate for current bandwidth condition, and keep the video quality both smooth and at a high level.

In this paper, an efficient adaptive video streaming system based on SVC is proposed with the above two problems better solved. Apple Darwin Streaming Server (DSS) \cite{DSS}, which is open-sourced and has been widely used to stream multimedia contents in practice, is chosen as the basis of the proposed system. The original DSS project is modified and SVC support is added to it, for the purpose of both research and application. Then novel algorithms for bitstream extraction and quality control are designed and implemented to achieve significant improvement in terms of user-perceived video quality. The contributions of our work can be summarized as follows:
\begin{itemize}
\item First, a simple and effective linear error model (LEM) is proposed and validated, which can be used to accurately estimate the distortion caused by discarding any combination of packets from a SVC bitstream.
\item Second, on the basis of the distortion estimation enabled by LEM, a greedy-like algorithm is designed to assign priority for each packet according to its R-D impact, thus enables optimized bitstream extraction.
\item Finally, for the quality control problem, a novel algorithm based on the Proportional-Integral-Derivative (PID) control concept is proposed to replace the native one in DSS, and achieves much lower quality fluctuation with higher quality level.
\end{itemize}

The rest of this paper is organized as follows. In section \ref{sec:analysis}, we analyze the problems in detail and introduce some related works on them. Then in section \ref{sec:bitstream-extraction}, the linear error model is proposed, and the solution for the bitstream extraction problem based on the model is present. After that, the PID-based quality control algorithm is proposed in section \ref{sec:quality-control}. Section \ref{sec:experiment} lists the experimental results and compares our proposed solution with others. Finally, in section \ref{sec:conclusion}, we conclude this paper's work and point out its promising applications and extensions in the future.

@@

In this section, we analyze the two most concerned problems in SVC adaptive video streaming, i.e., bitstream extraction and quality control, and introduce some existing related works on them.

@@

The following of this paper will present our solutions for the above problems in detail. And we will combine our works together and lay a solid foundation for a practical adaptive video streaming system.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Optimized Bitstream Extraction Based on Linear Error Model}
\label{sec:bitstream-extraction}

Being able to adjust bitrate is the prerequisite for an adaptive video streaming system. For SVC videos, it means to extract a subset of packets from the whole bitstream. How to optimally choose this subset to achieve minimum distortion under the given bitrate is the key point of bitstream extraction problem. In this section, we propose a linear error mode and a greedy-like algorithm to solve this problem.

@@

@@

To implement our proposed adaptive streaming system requires the combination of the optimized bitstream extraction method and the PID-based quality control algorithm. The latter determines the current bitrate to follow the bandwidth change, and the former will ensure the sub-stream with the best possible quality is obtained and sent under that bitrate constraint. They work together to construct a system that delivers best quality of experience for the users. Choosing DSS as the implementation basis, we have integrated the proposed algorithms into it for the purpose of both research and application. In this section, we first present separate experiments to demonstrate the advantage of each proposed solution, and then give the result when they are combined together. Readers who are interested in the practical implementation can also visit the online video website http://www.7dlive.com for more information.
